Towards Generic Parallel Programming in Computer Science Education with Kokkos

### 2020-08-10 src/0.Abstract.tex
Fix typos highlighted by the reviewers.

\begin{abstract}\label{chap:abstract}
Kokkos is a generic, performance-portable parallel programming model for host- and accelerator architectures in C++. It consists of a pure C++ interface, a specification and a programming library. The programming library exposes patterns and types and maps them to an underlying abstract machine model. 
This machine model offers a generic view of parallel hardware by abstracting from architecture- and vendor-specific programming models. In academia, this can contribute to a generic eduction of performance-portable parallel programming. In this work we present Kokkos from the perspective of educators, highlight key concepts and give an instruction to examples suited for incremental teaching of parallel programming in class.
\end{abstract}

%is aligned with C++ standardization efforts and trends towards parallel programming support in base languages.

>>>>

\begin{abstract}\label{chap:abstract}
Parallel patterns, views, and spaces are promising abstractions to capture the programmer's intent as well as the contextual information that can be used by an underlying runtime to efficiently map software to parallel hardware. These abstractions can be valuable in cases where an algorithm must accommodate requirements of code and performance portability across hardware architectures and vendor programming models. Kokkos is a parallel programming model for host- and accelerator architectures that relies on these abstractions and targets these requirements. It consists of a pure C++ interface, a specification, and a programming library. The programming library exposes patterns and types and maps them to an underlying abstract machine model. The abstract machine model offers a generic view of parallel hardware. While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to parallel programming productivity. In this work, we give an insight into design considerations of a representative programming model, list important abstractions, document feedback received during workshops, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
\end{abstract}

### 2020-08-10 src/1.Introduction.tex
Fix typos highlighted by the reviewers.

\section{Introduction}\label{chap:introduction}

@@

While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined. In this case, educators are exposed to a variety of solutions, approaches and trends in parallel computing. Which particular set to choose for teaching in class is not trivial. While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming generically is among the most important ones.

Education of parallel programming today is difficult. Educators strive to balance considerations of architecture, programming model, and theoretical concepts. Unfortunately, not all architectures support all programming models, nor do many programming models offer enough abstractions to support modern hardware architectures sufficiently. Many vendor programming models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine specification. We would like to compare this to the support of Von Neumann architectures in programming languages today.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively by a number of national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely execution patterns and data structures which it maps to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}. We list its key properties as follows.

\begin{itemize}
\item Parallel programming model for C++
\item Targets an abstract machine model
\item Implements parallel patterns and tasking paradigms
\item Offers abstract data types for performance portability
\item Pure C++ interface through template metaprogramming
\item Aligns with C++ standardization efforts for parallel programming
\item Vendor- and architecture independent and open-source
\item Includes exercises and teaching material
\end{itemize}

The objective of this paper is to present Kokkos from three perspectives. Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model. These insights may be  useful in parallel programming classes that introduce programming models and reason about design choices.

Further, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative. This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.

Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions. We show a set of interfaces that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

The rest of this paper is structured as follows. The next Chapter gives an overview on parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application. In Chapter~\ref{chap:backend} we provide insight into a Kokkos back-end implementation. Chapters~\ref{chap:related} discusses related work and lastly Chapter~\ref{chap:conclusion} concludes this work.

>>>>

@@

While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined. In this case, educators can select from a variety of solutions, approaches, and trends in parallel computing. Which particular set to choose for teaching in class is not trivial. Complexity is further amplified by the breath of related topics ranging from parallel hardware architectures, programming models, algorithms as well as productivity in parallel software development. However, it is the productivity concern that is gaining importance as the number of hardware solutions, their complexity as well as demands for a quick time-to-solution increase.

While vendor programming models maintain their position in curricula today, we think that the inclusion of parallel programming models such as Kokkos is worth consideration. They offer an easy to understand set of concepts for a vendor- and architecture-independent, performance-portable expression of concurrency which significantly contributes to parallel programming productivity.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS} is a performance-portable, parallel programming model for C++ developed collaboratively by several national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}. It consists of a programming model specification and a library. The specification defines the execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely execution patterns and types, including views and execution spaces and maps them to the underlying abstract machine model. The machine model creates a generic view of hardware and memories. Finally, the Kokkos library implements the mapping of the programming model to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at Sandia National Laboratories, Oak-Ridge National Laboratories, Los Alamos National Laboratories, University of Utah, and others~\cite{KOKKOSPROJS}. We list its key properties as follows.

\begin{itemize}
\item Targets an abstract machine model
\item Implements patterns, views, and spaces
\item Vendor- and architecture-independent and open-source
\item Aligns with C++ standardization efforts for parallel programming
\item Includes exercises and teaching material
\end{itemize}

The objective of this paper is to present Kokkos from three perspectives. Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model. These insights may be useful in parallel programming classes that introduce programming models, reason about design choices, and cover productivity in parallel software development.

Further, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a way that is generic, vendor-agnostic, and declarative. Abstractions allow students to learn about concurrency without being exposed to many architectural concerns initially and help to create awareness of their productivity benefits in terms of code- and performance portability on current and upcoming hardware architectures.

Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions in parallel programming models. We show a set of interfaces that implement patterns, views, and spaces, and capture enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++. For teaching insights, we document feedback received during on-line lectures.

The rest of this paper is structured as follows. The next chapter gives an overview of parallel programming concepts and presents the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive the required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application. In Chapter~\ref{chap:backend} we provide insight into a Kokkos back-end implementation. Chapters~\ref{chap:related} and ~\ref{chap:tutorials} discuss related work, tutorials and student feedback from tutorials. Lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-08-10 src/2.GenericProgramming.tex
Fix typos highlighted by the reviewers.

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often through \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

@@

\begin{figure}[h]
\begin{minted}[fontsize=\small]{text}
Constrained Semantics: memory allocations, 
memory movement, execution parameters
Parallel Paradigms: patterns, tasks
Expression: language 
Control Paradigm: prescriptive, descriptive
\end{minted}
\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
\label{figSemCapture}
\end{figure}

@@

A \emph{Parallel Paradigm} is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to choose depends on several considerations. Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels. This paradigm eliminates accesses to shared state and aligns well with message passing programming (MPI). The execution model and memory model aspects of a programming model define its behavior. That is, it defines the relationship between abstract concepts and program execution on the given architecture.

A language defines the syntax of the parallel semantic. Annotations-based  languages add semantic information to the base language through pragmas while embedded programming models rely on the base language and offer additional programming interfaces to capture information. Their semantic is defined in the API specification.

Lastly, a programming model must divide responsibility between the developer and the model. We refer to is as defining the \emph{Control Paradigm}. In practice, this is a choice between defining a descriptive and prescriptive semantics. Descriptive semantics require a developer to express intent. The implementation is the responsibility of the programming model and of the underlying toolchain. Prescriptive semantics require the developer to specify the mechanisms by which their intent will be implemented. In a prescriptive model, developer success is tied to their knowledge of implementation mechanisms. In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into an implementation.

Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show two code examples using the aforementioned parallel programming syntaxes. While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations. Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantic of describing a parallel loop construct. It is up to the programming model specification to make further definitions regarding separation of concern.

Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures efficiently. In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first. In the following chapters we use the words \emph{representation} and \emph{model} interchangeably. In this work the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.

>>>>

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often through \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with a sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

@@

\begin{figure}[h]
\begin{minted}[fontsize=\small]{text}
Constrained Semantics: memory allocations, 
memory movement, execution parameters
Parallel Paradigms: patterns, tasks
Expression: language 
Control Paradigm: prescriptive, descriptive
\end{minted}
\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties of a parallel programming model.}
\label{figSemCapture}
\end{figure}

@@

A \emph{Parallel Paradigm} is an abstract representation to facilitate the programmer's understanding of programming rules and program behavior. Which programming paradigm to choose depends on several considerations. Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels. This paradigm eliminates access to a shared state and aligns well with message passing programming (MPI). The execution model and memory model aspects of a programming model define its behavior. That is, it defines the relationship between abstract concepts and program execution on the given architecture.

A language defines the syntax of the parallel semantics. Annotations-based languages add semantic information to the base language through pragmas while embedded programming models rely on the base language and offer additional programming interfaces to capture information. Their semantics is defined in the API specification.

Lastly, a programming model must divide responsibility between the developer and the model. We refer to this as defining the \emph{Control Paradigm}. In practice, this is a choice between defining a descriptive and prescriptive semantics. Descriptive semantics requires a developer to express intent. The implementation is the responsibility of the programming model and of the underlying toolchain. Prescriptive semantics requires the developer to specify the mechanisms by which their intent will be implemented. In a prescriptive model, developer success is tied to their knowledge of implementation mechanisms. In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into implementation.

Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show two code examples using the aforementioned parallel programming syntaxes. While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations. Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantics of describing a parallel loop construct. It is up to the programming model specification to make further definitions regarding the separation of concern.

Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow them to map parallel execution on modern architectures efficiently. To define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first. In the following chapters we use the words \emph{representation} and \emph{model} interchangeably. In this work, the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.

### 2020-08-10 src/3.MachineModel.tex
Fix a few typos
Fix typos highlighted by the reviewers.

The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access. The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API. By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware. The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model. We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristic and portability remain the same.

The Kokkos abstract machine representation targets a design of a future shared-memory computing architectures. The design is shown in Figure~\ref{fig:spaces}. It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators. Both compute device types can have disjoint memory address spaces with unique performance properties. In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches. In order to ensure performance portability on such a wide range of configurations, an abstraction of compute resources and available memories are required. For this purpose we introduce the abstract concept called \emph{space}.

@@

An instance of a memory space provides a concrete method for the application programmer to request data storage allocations. Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM and non-volatile memories. Accelerators can provide an additional memory space through its local on-package memory. This is also highligted in Figure~\ref{fig:spaces}. The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space. Programmatically this is achieved by instantiating the according memory space. Kokkos provides the appropriate abstraction of memory allocation and data management. We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.

>>>>

The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access. The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API. By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware. The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model. We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristics and portability remain the same.

The Kokkos abstract machine representation targets a design of a future shared-memory computing architecture. The design is shown in Figure~\ref{fig:spaces}. It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators. Both compute device types can have disjoint memory address spaces with unique performance properties. In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches. In order to ensure performance portability on such a wide range of configurations, an abstraction of compute-resources and available memories is required. For this purpose, we introduce the abstract concept called \emph{space}.

@@

An instance of a memory space provides a concrete method for the application programmer to request data storage allocations. Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM, and non-volatile memories. Accelerators can provide an additional memory space through its local on-package memory. This is also highlighted in Figure~\ref{fig:spaces}. The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space. Programmatically this is achieved by instantiating the according memory space. Kokkos provides the appropriate abstraction of memory allocation and data management. We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.

### 2020-08-10 src/4.ProgammingModel.tex
Fix typos highlighted by the reviewers.

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in industry in general express the demand for C++ and a generic parallel programming model to support their codes. Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing. Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.

From the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries. In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.

@@

To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}. These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections. 
	\item Execution policies shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from array indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access and streaming loads or stores. This allows the programming model to  optimize load and store operations.
\end{itemize}

@@

Figure~\ref{fig:parallelFor} shows one definition of the Kokkos~\emph{parallel\_for} interface. This interface accepts two template parameters and three function arguments. \emph{ExecPolicy} shapes the iteration space for a particular execution space. Examples of execution policies are the~\emph{RangePolicy}, ~\emph{MDRangePolicy} and the~\emph{TeamPolicy}. As the name suggests, these policies express iteration spaces. For example, a range policy defines a one dimensional iteration range. The MDRangePolicy represents a multi-dimensional iteration space and is used to express tightly nested loop patterns. A team policy defines a 1D iteration range, each of which is assigned to a team of threads. This policy allows the expression of hierarchical parallelism. The second template parameter,~\emph{FuntorType}, represent a functor that implements the~\emph{()-operator} with a matching signature for the given execution policy. The functor can be defined using a C++ class, struct or lambda expression. We would like to highlight that the return semantic of the parallel\_for is defined as \emph{potentially asynchronous}. To guarantee a kernel has finished, a developer should call the fence of the execution space on which the kernel is being executed. Otherwise it depends on the execution space where the loop executes and weather this execution space implements a barrier.

The semantic of the parallel\_for class in Kokkos demonstrates the descriptive nature of the programming model. Programming with patterns and abstractions allows the compiler to generate optimized transformations. 
Generally parallel patterns in Kokkos do not guarantee iteration ordering nor degree of concurrency during their execution. This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes. We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for generic support for parallel programming. In conclusion, Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.

To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.
The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the the underlying back-end programming model.

>>>>

Kokkos defines a pure interface for C++ and uses C++ as a programming language. This has multiple reasons. It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in the industry in general express the demand for C++ and a generic parallel programming model to support their codes. Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represents a burden worth addressing. Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.

From the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries. In this case, class specialization and templating allow for compile-time generated types and optimizations for a given hardware architecture.

@@

To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}. These abstractions specify semantic information that enables to capture the programmer's intent and allows the runtime to efficiently map the program to any underlying hardware architecture. We list them as follows:
\begin{itemize}
	\item  An execution space is a place where code can be executed. On current hardware architectures, this corresponds to accelerators and CPUs and can include other compute devices in the future. This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
	\item Execution patterns expose the parallel programming paradigm. Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections. 
	\item Execution policies shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
	\item Memory spaces are the places where data resides. They specify the physical locations of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories, or non-volatile bulk storage. Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
	\item Layouts express the mapping from array indices to address offsets. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
	\item Memory traits specify how a data structure is accessed. Traits express usage scenarios such as atomic access, random access, and streaming loads or stores. This allows the programming model to optimize load and store operations.
\end{itemize}

@@

Figure~\ref{fig:parallelFor} shows one definition of the Kokkos~\emph{parallel\_for} interface. This interface accepts two template parameters and three function arguments. \emph{ExecPolicy} shapes the iteration space for a particular execution space. Examples of execution policies are the~\emph{RangePolicy}, ~\emph{MDRangePolicy} and the~\emph{TeamPolicy}. As the name suggests, these policies express iteration spaces. For example, a range policy defines a one-dimensional iteration range. The MDRangePolicy represents a multi-dimensional iteration space and is used to express tightly nested loop patterns. A team policy defines a 1D iteration range, each of which is assigned to a team of threads. This policy allows the expression of hierarchical parallelism. The second template parameter,~\emph{FunctorType}, represent a functor that implements the~\emph{()-operator} with a matching signature for the given execution policy. The functor can be defined using a C++ class, struct or lambda expression. We would like to highlight that the return semantic of the parallel\_for is defined as \emph{potentially asynchronous}. To guarantee a kernel has finished, a developer should call the fence of the execution space on which the kernel is being executed. Otherwise, it depends on the execution space where the loop executes and whether this execution space implements a barrier.

The semantic of the parallel\_for class in Kokkos demonstrates the descriptive nature of the programming model. Programming with patterns and abstractions allows the compiler to generate optimized transformations. 
Generally, parallel patterns in Kokkos do not guarantee iteration ordering nor the degree of concurrency during their execution. This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as the assignment of iterations to threads or vector lanes. We believe that these abstractions and their descriptive semantic represent key programming primitives and the way forward for generic support for parallel programming. In conclusion, Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.

To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.
The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the underlying back-end programming model.

### 2020-08-10 src/5.Example.tex
Fix typos highlighted by the reviewers.

Views are abstractions that represent data. Naturally data can reside in host memory where it is accessible to the host execution space or on device memory where it is accessible to the device execution space. In this example, copies of the views are created in order to make them accessible by the host process. This is accomplished by calling the ~\emph{create\_mirror\_view} function.

If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation. In real applications host-accessible allocation is necessary in order to allow I/O operations during filesystem access. To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked. In case two view objects are aliased, this operation results in a~\emph{no-op}.

%A general philosophy of Kokkos is to have defaults for most properties, including data layouts, memory spaces and execution spaces.
Consequently the simple example does not explicitly set which memory space the data lives in.
Kokkos ensures that the defaults for parallel execution and data allocations match though.

Taking a closer look at the loop step shows the hierarchical expression of concurrency. In case of the matrix-vector multiplication,This is accomplished by using the ~\emph{Kokkos::TeamPolicy}. For each row of~\emph{A} a team of threads is launched, where the size of the team is not further specified. When using a ~\emph{TeamPolicy} the operator of the lambda expression does not receive an index, but rather a handle to the team of threads. 
This handle provides the team identifier \emph{e}, and is subsequently handed to the nested parallel execution policies. The nested reduction requires the operator of the nested lambda to take a reference to the thread-local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index. It then writes the result to the correct position of the output vector. 
The parallel loops are follows by a call to \emph{Kokkos::fence}. This ensures that the execution of the parallel code is completed before copying the results back to the host memory space. This is implemented through the~\emph{Kokkos::deep\_copy} call. Lastly, the Kokkos application calls~\emph{Kokkos::finalize}.

While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources. Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide and an API documentation. Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}. The tutorial is intended for students with minimal or even no prior knowledge of parallel programming. It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises which build up on each other and gradually introduce new concepts. The programming guide is available on-line~\cite{KOKKOS_WIKI}.

>>>>

Views are abstractions that represent data. Naturally, data can reside in host memory where it is accessible to the host execution space or on device memory where it is accessible to the device execution space. In this example, copies of the views are created to make them accessible by the host process. This is accomplished by calling the ~\emph{create\_mirror\_view} function.

If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation. In real applications, a host-accessible allocation is necessary to allow I/O operations during filesystem access. To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked. In case two view objects are aliased, this operation results in a~\emph{no-op}.

Consequently, the simple example does not explicitly specify a memory space where the data lives in. However, Kokkos ensures that the defaults for parallel execution and data allocations match.

Taking a closer look at the loop step shows the hierarchical expression of concurrency. This is accomplished by using the ~\emph{Kokkos::TeamPolicy}. For each row of~\emph{A} a team of threads is launched, where the size of the team is not further specified. When using a ~\emph{TeamPolicy} the operator of the lambda expression does not receive an index, but rather a handle to the team of threads. 
This handle provides the team identifier \emph{e}, and is subsequently handed to the nested parallel execution policies. The nested reduction requires the operator of the nested lambda to take a reference to the thread-local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index. It then writes the result to the correct position of the output vector. 
The parallel loops are followed by a call to \emph{Kokkos::fence}. This ensures that the execution of the parallel code is completed before copying the results back to the host memory space. This is implemented through the~\emph{Kokkos::deep\_copy} call. Lastly, the Kokkos application calls~\emph{Kokkos::finalize}.

While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources. Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation. Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}. The tutorial is intended for students with minimal or even no prior knowledge of parallel programming. It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises that build upon each other and gradually introduce new concepts. The programming guide is available on-line~\cite{KOKKOS_WIKI}.
